Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-15 aistats.tex

\subsection{Asymptote}
\label{ssec:asymptote}
As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
We now detail a classical asymptotic result.
Proofs are in App.\ref{app:asymptote}, and \citet[\S1.1]{ostrovskii2021finite} offers a great review.

Statistics typically give results on $\nat$, but the MAP~\eqref{eq:defMAP} is more simply expressed with $\meanp$, so let us focus on $\bregmanconj$.
Bregman divergences are locally quadratic, as seen via a second order Taylor expansion
\alignn{
    \textstyle \bregmanconj(\m^* ; \m)
    &\textstyle = \frac{1}{2}\norm{\m^* - \m}^2_{\mF}
    + O(\norm{\m - \m^*}^3),
    \label{eq:bregmanTaylor}
}
where the Mahalanobis norm  $\| x \|_{\mF}^2 = x^\top \mF x$  is induced by $\mF  := \nabla^2\conj(\m^*)$, the hessian of the entropy at the optimum. It happens that  $\mF$ is also the inverse \textit{Fisher information matrix} at $\nat^*$, since
\aligns{
    \mF
    :=\nabla^2\conj(\m^*)
    = \nabla^2\logpart(\nat^*)^{-1}
    = \Cov_{\nat^*}[T(X)]^{-1}  \; .
}
Plugging the MLE~\eqref{eq:defMLE} or MAP~\eqref{eq:defMAP} into~\eqref{eq:bregmanTaylor}, we have
\begin{align}
	\label{eq:asymptote}
	\E \bregmanconj \left (\E [T(X)] ; \hat \meanp_n^\text{MLE/MAP} \right )
	= \frac{d}{2n} + O(n^{- \frac{3}{2}}) \; .
\end{align}
Both MLE and MAP have the same asymptote, as the contribution of the prior $n_0 \meanp_0$ gets negligible for large $n$.
This asymptote is independent of the optimum $\meanp^*$ or $\mF$ for well-specified models.
Let us see another example for which we get a rate matching~\eqref{eq:asymptote}.

\subsection{Quadratic Case}
\label{ssec:quadratic}
As another classical reference point, we consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
For instance, this is the log-partition of a Gaussian with known variance $I$,
\[
	\cX=\real^d,\quad \nu(dx) = \exp\paren{\textstyle \half[-\|x\|^2]} dx,\quad T(x)=x.
\]
In this case, $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances since
\begin{align}
	\bregmanconj(\meanp^* ; \meanp) = \half \norm{\meanp^* -  \meanp }_2^2  \; .
\end{align}
Thanks to the independence of samples, we can break down the MLE into individual point's contributions:
\begin{align}
	\expect{\half \norm{\m^* -  \inv{n}  \smallsum_i T_i}_2^2}
	=\frac{\Var(T)}{2n}
	=\frac{d}{2n}.
\end{align}
Adding a reference mean $\m_0$ to get the MAP yields
\begin{align}
	\expect{\bregmanconj(\meanp^*; \MAPm)}
	&= \frac{n \Var(T) +  n_0^2 \norm{\m^* -  \m_0}^2}{2(n+n_0)^2}.
	%\\ &= O\left(\frac{\Var(T)}{n} \right) + O\left(\frac{\norm{\m^* -  \m_0}^2}{n^2} \right)
	\label{eq:MAP_quadratic}
\end{align}
We see here a variance term defining the $\frac{d}{2n}$ asymptote and a bias term in $O(n^{-2})$. However, this result does not generalize well to other families unless we make restrictive assumptions on $\conj$. % we can relate $\bregmanconj$ to a norm or a quadratic.
% Can we generalize these results to other families?

@@

\subsection{Locally Quadratic Case}
\label{ssec:local-quadratic}
From Taylor expansion~\eqref{eq:bregmanTaylor},
we know that all Bregman divergences are locally quadratic.
Under some assumptions, such as self-concordance\footnote{
In 1d, $f$ is self-concordant iff $\forall x, \abs{f'''(x)} \leq 2 \abs{f''(x)}^{\frac{3}{2}}$.
} of $\conj$ \citep[Ch.4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in. Proofs for this subsection are in App.\ref{app:self-concordant}.
\begin{proposition}
Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$. Then\footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}
\aligns{
	\norm{\m^*-\m}_{\mF} < 0.21
	\implies
	\bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF}^2.
}
\end{proposition}
To gain insights into how many samples are needed, we can estimate when $\expecti{\norm{\m^*-\m}_{\mF}} < 0.21 $.
For the MLE, $\expecti{\norm{\m^*-\hat \m_n}_{\mF}} = \frac{d}{n}$, so a sufficient condition is $n \geq 5 d$.
For MAP, using~\eqref{eq:MAP_quadratic}, we get the sufficient condition $n\geq 5d + \sqrt{5}\norm{\m^* -  \m_0} - n_0$.
This means that, on average, we need $5$ times more samples than the dimension to reach the quadratic regime and ensure an upper bound like~\eqref{eq:MAP_quadratic}.

The entropy $\conj$ is self-concordant for a lot of common families.
For instance, when $T$ is 1-dimensional, and $\logpart$ is self-concordant (App.\ref{app:self-concordant}),
as is the case with:
exponential distributions,
Gaussians with known mean,
Laplace with known mean,
Pareto with known minimum value,
or Weibull with known shape $k$.
The entropy is also self-concordant when $T$ lives in a compact \citep{bubeck2015entropic} -- e.g., categorical and Dirichlet distributions.

@@

These results partially answer~\eqref{problem}.
Partially because the present proposition and these related works all give \textit{large sample} results,  that hold when $n\geq N$ for some constant $N$.
None of them apply to small $n$.
That is one caveat. Another one is the gaussian example we saw in \S\ref{ssec:gaussian}, which is not self-concordant.
A full solution to~\eqref{problem} would address both of these caveats.
Informed by the properties we have seen so far, let us investigate a general decomposition of the Bregman that could guide us towards a solution.
%If no current approach solves~\eqref{problem}, then let us formulate a conjecture informed by the examples and partial solutions we have s.

>>>>

\subsection{Asymptotic Rate}
\label{ssec:asymptote}
As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
We now detail a classical asymptotic result.
Proofs are in App.~\ref{app:asymptote}, and \citet[\S1.1]{ostrovskii2021finite} offers a great review.

Statistics typically give results on $\nat$, but the MAP~\eqref{eq:defMAP} is more simply expressed with $\meanp$, so let us focus on $\bregmanconj$.
Bregman divergences are locally quadratic, as seen via a second order Taylor expansion
\alignn{
    \textstyle \bregmanconj(\m^* ; \m)
    &\textstyle = \frac{1}{2}\norm{\m^* - \m}^2_{\mF}
    + O(\norm{\m - \m^*}^3),
    \label{eq:bregmanTaylor}
}
where the Mahalanobis norm  $\| x \|_{\mF}^2 = x^\top \mF x$  is induced by $\mF  := \nabla^2\conj(\m^*)$, the Hessian of the entropy at the optimum. It happens that  $\mF$ is also the inverse \textit{Fisher information matrix} at $\nat^*$, since
\aligns{
    \mF
    :=\nabla^2\conj(\m^*)
    = \nabla^2\logpart(\nat^*)^{-1}
    = \Cov_{\nat^*}[T(X)]^{-1}  \; .
}
Plugging the MLE~\eqref{eq:defMLE} or MAP~\eqref{eq:defMAP} into~\eqref{eq:bregmanTaylor}, we have
\begin{align}
	\label{eq:asymptote}
	\E \bregmanconj \left (\E [T(X)] ; \hat \meanp_n^\text{MLE/MAP} \right )
	= \frac{d}{2n} + O(n^{- \frac{3}{2}}) \; .
\end{align}
Both MLE and MAP have the same asymptote, as the contribution of the prior $n_0 \meanp_0$ gets negligible for large $n$.
This asymptote is independent of the optimum $\meanp^*$ or $\mF$ for well-specified models.
Next, we give another example for which we get a rate matching~\eqref{eq:asymptote}.

\subsection{Quadratic Case}
\label{ssec:quadratic}
As another classical reference point, we consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
For instance, this is the log-partition of a Gaussian with known variance $I$,
\[
	\cX=\real^d,\quad \nu(dx) = \exp\paren{\textstyle \half[-\|x\|^2]} dx,\quad T(x)=x.
\]
In this case, $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances since
\begin{align}
	\bregmanconj(\meanp^* ; \meanp) = \half \norm{\meanp^* -  \meanp }_2^2  \; .
\end{align}
Thanks to the independence of samples, we can break down the MLE into individual point's contributions:
\begin{align}
	\expect{\half \norm{\m^* -  \inv{n}  \smallsum_i T_i}_2^2}
	=\frac{\Var(T)}{2n}
	=\frac{d}{2n}.
\end{align}
Adding a reference mean $\m_0$ to get the MAP yields
\begin{align}
		\!\!\!\!\! \expect{\half \norm{\m^* -   \MAPm^\text{MAP}}_2^2} \!
	&= \frac{n \Var(T) +  n_0^2 \norm{\m^* -  \m_0}^2}{2(n+n_0)^2}.\!
	%\\ &= O\left(\frac{\Var(T)}{n} \right) + O\left(\frac{\norm{\m^* -  \m_0}^2}{n^2} \right)
	\label{eq:MAP_quadratic}
\end{align}
We see here a variance term defining the $\frac{d}{2n}$ asymptote and a bias term in $O(n^{-2})$. However, this result does not generalize well to other families unless we make restrictive assumptions on $\conj$. % we can relate $\bregmanconj$ to a norm or a quadratic.
% Can we generalize these results to other families?

@@

\subsection{Locally Quadratic Case}
\label{ssec:local-quadratic}
From Taylor expansion~\eqref{eq:bregmanTaylor},
we know that all Bregman divergences are locally quadratic.
Under some assumptions, such as self-concordance\footnote{
In 1d, $f$ is self-concordant iff $\forall x, \abs{f'''(x)} \leq 2 \abs{f''(x)}^{\frac{3}{2}}$.
} of $\conj$ \citep[Ch.~4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in. Proofs for this subsection are in App.~\ref{app:self-concordant}.
\begin{proposition}
Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$. Then\footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$.}
\aligns{
	\norm{\m^*-\m}_{\mF} < 0.21
	\implies
	\bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF}^2.
}
\end{proposition}
To gain insights into how many samples are needed, we can estimate when $\expecti{\norm{\m^*-\m}_{\mF}} < 0.21 $.
For the MLE, the proof of~\eqref{eq:asymptote} from~\eqref{eq:bregmanTaylor} yields $\expecti{\norm{\m^*-\hat \m_n}^2_{\mF}} = \frac{d}{n}$ in general, so a sufficient condition is $n \geq 25 d$.
For MAP, transforming~\eqref{eq:MAP_quadratic}, we get the sufficient condition \TODO{FIX THIS!} $n\geq 5d + \sqrt{5}\norm{\m^* -  \m_0} - n_0$.
This means that on average, we need $25$ times more samples than the dimension to reach the quadratic regime and ensure an upper-bound like~\eqref{eq:MAP_quadratic}.

The entropy $\conj$ is self-concordant for several common families.
For instance when $T$ is 1-dimensional and $\logpart$ is self-concordant (App.~\ref{app:self-concordant}),
as is the case with:
exponential distributions,
Gaussians with known mean,
Laplace with known mean,
Pareto with known minimum value,
or Weibull with known shape $k$.
The entropy is also self-concordant when $T$ lives in a compact \citep{bubeck2015entropic} -- e.g., categorical and Dirichlet distributions.

@@

These results partially answer~\eqref{problem}.
Partially because the present proposition and these related works all give \textit{large sample} results,  that hold when $n\geq N$ for some constant $N$.
None of them apply to small $n$.
That is one caveat. Another one is the Gaussian example we saw in \S\ref{ssec:gaussian}, which is not self-concordant.
A full solution to~\eqref{problem} would address both of these caveats.
Informed by the properties we have seen so far, let us investigate a general decomposition of the Bregman that could guide us towards a solution.
%If no current approach solves~\eqref{problem}, then let us formulate a conjecture informed by the examples and partial solutions we have s.
